---
title: "BoundarySpatial_in_suite.R Demo"
subtitle: "Spatial Transcriptomics Tumor Boundary Extraction & Distance Belt Profiling"
author: "Changxiang Huan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk $ set(
  echo      = TRUE,
  warning   = FALSE,
  message   = TRUE,
  fig.width = 14,
  fig.height = 6,
  dpi       = 300
)

```

# 1. Environment Setup

Load the `BoundarySpatial_in_suite.R` script to import all core functions.

```{r load-script}
source("/data1/huanchangxiang/Python_script/小论文一/BoundarySpatial_in_suite.R")

```

# 2. Data Loading & Preview

```{r load-data}
input_path <- "/data1/huanchangxiang/datasets/public_spatial/output/02Xenium/Breast_alpha_shape.csv"
df <- read.csv(input_path)

cat("Data dimensions:", nrow(df), "spots ×", ncol(df), "columns\n")
cat("Column names:", paste(colnames(df), collapse = ", "), "\n\n")
cat("Cluster distribution:\n")
print(table(df $ clusters))

```

```{r quick-overview, fig.width=8, fig.height=6}
ggplot(df, aes(x = x, y = y, color = as.factor(clusters))) +
  geom_point(size = 0.6, alpha = 0.7) +
  scale_color_manual(values = c("0" = "grey30", "1" = "#FF0000",
                                "2" = "#B29CDC", "3" = "#B1D685")) +
  theme_minimal() +
  labs(title = "Spatial Overview — All Clusters",
       color = "Cluster") +
  coord_fixed()

```

# 3. Tumor Boundary Polygon Extraction
## 3.1 extract_tumor_boundaries
Call `extract_tumor_boundaries()` to select Tumor core region clusters. `eps`: Density radius for DBSCAN. `min_pts`: It is recommended to set a value greater than 50 to identify the primary tumor core regions. For capturing more detailed boundaries, this parameter should be tuned iteratively.

```{r step1-boundary-extraction}
boundaries <- extract_tumor_boundaries(
  data            = df,
  target_clusters = c("0","1"),
  mode            = "multi",
  eps             = 100,
  min_pts         = 100,
  concavity       = 1.2
)

```

## 3.2 Boundary Visualization

```{r step1-plot-boundaries, fig.width=10, fig.height=8}
ggplot() +
  geom_point(data = df, aes(x = x, y = y, color = as.factor(clusters)),
             size = 0.5, alpha = 0.6) +
  geom_sf(data = boundaries, fill = NA, color = "white",
          linewidth = 1, linetype = "dashed", inherit.aes = FALSE) +
  scale_color_manual(values = c("0" = "grey30", "1" = "#FF0000",
                                "2" = "#B29CDC", "3" = "#B1D685")) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "black", color = NA),
        plot.background  = element_rect(fill = "black", color = NA),
        legend.text      = element_text(color = "white"),
        legend.title     = element_text(color = "white"),
        legend.key       = element_rect(fill = "black"),
        axis.text        = element_text(color = "white"),
        panel.grid.major = element_line(color = "grey20", linewidth = 0.2)) +
  labs(title = "Tumor Boundary Extraction",
       color = "Cluster") +
  coord_sf()
```

# 4. Distance Calculation & Spatial Belt Classification

Call `calculate_distance_belts()` to compute the shortest Euclidean distance from each spot to the tumor boundary, and classify them into distance belts with a 100μm step size.

```{r step2-distance-calculation}
df_final <- calculate_distance_belts(
  df         = df,
  boundaries = boundaries,
  bin_size   = 100,
  max_dist   = 1000
)

```

## 4.1 Distance Distribution Overview

```{r step2-distance-summary}
cat("New columns:", paste(setdiff(colnames(df_final), colnames(df)), collapse = ", "), "\n\n")
cat("Spot count per distance belt:\n")
print(table(df_final $ dist_group))
cat("\nDistance statistics (μm):\n")
cat(sprintf("  Min distance:    %.1f\n", min(df_final $ dist_to_border)))
cat(sprintf("  Median distance: %.1f\n", median(df_final $ dist_to_border)))
cat(sprintf("  Max distance:    %.1f\n", max(df_final $ dist_to_border)))
cat(sprintf("  Mean distance:   %.1f\n", mean(df_final $ dist_to_border)))

```

## 4.2 Distance Histogram

```{r step2-distance-hist, fig.width=10, fig.height=5}
ggplot(df_final, aes(x = dist_to_border)) +
  geom_histogram(binwidth = 50, fill = "#fdae61", color = "white", alpha = 0.9) +
  geom_vline(xintercept = 5, color = "#d53e4f", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 30, y = Inf, label = "Bdy cutoff (5μm)",
           color = "#d53e4f", hjust = 0, vjust = 1.5, size = 3.5) +
  theme_minimal() +
  labs(title = "Distribution of Distance to Tumor Boundary",
       x = "Distance to Boundary (μm)",
       y = "Spot Count")

```

# 5. Spatial Visualization

Call `plot_spatial_results()` to generate both the boundary extraction plot and the distance belt plot.

```{r step3-visualization}
my_palette <- c(
  'Tumor'  = '#FFC0CB',
  "Bdy"    = "#d53e4fff",
  "group0" = "#f46d43ff", "group1" = "#fdae61ff", "group2" = "#fee08b",
  "group3" = "#e6f598",   "group4" = "#abdda4ff", "group5" = "#66c2a5ff",
  "group6" = "#7ad6fb",   "group7" = "#7ebefb",   "group8" = "#b37ad1",
  "group9" = "#8849a7",   "other"  = "#BBDCAD"
)

plots <- plot_spatial_results(
  df         = df,
  boundaries = boundaries,
  df_final   = df_final,
  palette    = my_palette
)

```

## 5.1 Tumor Boundary Extraction Plot (BoundarySP)

```{r step3-plot-boundary, fig.width=14, fig.height=6}
plots $ boundary_plot
```

## 5.2 Distance Belt Spatial Distribution Plot (Boundary Both)

```{r step3-plot-belt, fig.width=14, fig.height=6}
plots $ belt_plot
```

# 6. Export Results

```{r step4-export}
output_dir <- "/data1/huanchangxiang/Python_script/小论文一/BoundarySpatial/Output/"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
ggsave(file.path(output_dir, "SFig6_Breast_BoundarySP.png"),
       plot = plots $ boundary_plot, width = 13.8, height = 6.1, dpi = 300)
ggsave(file.path(output_dir, "SFig6_Breast_Boundary_both.png"),
       plot = plots $ belt_plot, width = 13.8, height = 6.1, dpi = 300)
write.csv(df_final, file.path(output_dir, "Breast_spatial_with_distance.csv"),
          row.names = FALSE)
cat("Output files:\n")
cat("  1.", file.path(output_dir, "SFig6_Breast_BoundarySP.png"), "        — Boundary extraction plot\n")
cat("  2.", file.path(output_dir, "SFig6_Breast_Boundary_both.png"), "     — Distance belt spatial plot\n")
cat("  3.", file.path(output_dir, "Breast_spatial_with_distance.csv"), "   — Full dataset with distance annotations\n")
cat("\n[SUCCESS] Pipeline completed.\n")

```